Language models as probabilistic cognitive models

### 2023-11-18 03-paper/paper-LLM-probmod.tex
revision formal notation §5 & 6

The RSA model's predictions are defined at the condition-level.
An LLM, on the other hand, first and foremost makes predictions about each item.
As mentioned previously, the items of the reference game experiment differ in which levels of features (color, shape, texture) instantiate the structure of the task shown in Figure \ref{fig:ref-game}, as well as the order of presentation of objects and words.
We expect both human and LLM predictions to vary between different items: e.g., humans seem to have preferences for some features \citep[e.g.,][]{QingFranke2013:Variations-on-a}, maybe even leading to over-production of informationally irrelevant material \citep[e.g.,][]{DaviesKatsos2010:Over-informativ,Rubio-Fernandez2019:Overinformative,DegenHawkins2020:When-redundancy}; and machines may be susceptible to the presentation of the order of choice options. \mf{@PT references from the NLP literature on effects of prompt variation?}
It therefore becomes an empirical question of whether a standard item-level score from LLMs provides a good fit, if we aim at predicting the human data separately for each item, not as a condition-level average.

As described in Section~\ref{motivation}, the standard procedure for assessing LLM predictions in multi-choice tasks is to use a winner-takes-all (WTA) strategy that just picks the best option for each item.
But since this strategy has conceptual and empirical weaknesses, we define the item-level LLM predictions using the more general softmax-approach, which contains the WTA-based approach as a limiting case, repeated here from above with slightly updated notation:
%
\begin{align*}
P_{\text{item}}^{\text{SM}}\left ( y_{ki} \mid C; \alpha_{c} \right ) \propto \expo \left (\alpha_{c} \ \text{S}_{ki} \right)\,.
\end{align*}
%
Item-level scores $\text{S}_{ki}$, as defined in Section~\ref{motivation}, are obtained from GPT-3.5 \mf{which model exactly, when assessed, cite paper}.
% \footnote{
  % Numerically, we worked with \emph{average} log-probabilities. As in our experiment, all choice options had the same number of words, this is largely irrelevant, except for the effect of the $\alpha$ parameter in the interpretation condition. Using length-corrected log-probabilities is the same as using raw log-probabilities with an $\alpha$ parameter divided by the number of words. In this way, we make the $\alpha$ parameter more meaningfully comparable across conditions.}
The task description $x_{k}$ for item $I_{k}$ is a text supplied as prompt (see Appendix~\ref{sec:examples-items-llm} for an example).
The choice options are categorized as in the human experiment.\footnote{
  In the current set-up the response type ``distractor'' has two instantiations in the production condition. Since choice options are a single word in the production condition, for simplicity, we lump both of the distractor words together and treat them as a single option by taking the sum of the log-probabilities for both distractor words.}
Allowing for random errors with probability $\epsilon_{C}$, as before,:
%
\begin{align*}
  P_{\text{item}}^{\text{SM}}\left ( y_{ki} \mid C; \alpha_{c}, \epsilon_{c} \right )
  = \epsilon_{c} \  P_{\text{item}}^{\text{SM}}\left ( y_{ki} \mid C; \alpha_{c}, \epsilon_{c} \right ) \ \ + \ \ \nicefrac{\epsilon_{c}}{3}   \,.
\end{align*}
%
The resulting likelihood function for item-level data with item-level predictions is:
%
\begin{align*}
 P^{\text{SM}}_{\text{item}}(D_{c} \mid C; \alpha_{c}, \epsilon_{c}) = \prod_{k = 1}^{m} \text{Multinomial} \left ( D_{c}^{k}, \tuple{P^{\text{SM}}_{\text{item}}(y_{ki} \mid C;  \alpha_{c}, \epsilon_{c})}_{1 \le l \le 3} \right )\,.
\end{align*}
%
Priors on parameters are as for the RSA model described in Section~\ref{sec:model-pred-from}.

@@

Sampling-based approximations of Bayesian posterior predictive $p$-values for the by-item analysis are very low (see Table~\ref{tab:sumStats}), suggesting that the unaggregated item-level LLM predictions are inadequate predictors of the human data.
Figure~\ref{fig:item-level-obs-pred} shows that the item-level LLM scores predict variance which is not borne out by the human data.
Concretely, the plots show, for each condition and item, mean posterior estimates of the model's predicted probability of choosing the target option ($x$-axis), together with the observed proportion of target choices in the human data ($y$-axis).
There is ample variation in the model's predictions, especially visible in the production condition, owing to the fact that the item-level scores of the LLM sometimes clearly favor another option than the target choice.
So, the model itself predicts systematic variability at the item level.
The human data, too, show variability at the item-level, but there is no (visual) indication that the item-level variability predicted by the LLMs is borne out by the human data.
These results suggest that LLM-based probabilistic predictions may imply item-level variance that is not attested in the human data.
Put more strongly, a model that seeks to predict what human participants choose on a by-item level, based on the most obvious item-level score, is ruled out by experimental data.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.9\textwidth]{00-pics/measures-overview.png}
  \caption{
    Schematic overview over three different approaches of obtaining condition-level predictions by aggregating over item-level scores.
    The basic item-level score is average next-word log probability.
    This can be taken as-is into averaging (narrow-scope aggregation), or first transformed from log- into probability-space, either with optimization before (wide scope) or after averaging (intermediate scope).
    The notation for the condition-level options uses pseudo-code to more clearly reflect the relevant operator scope differences (where triangles represent the pipe operator).
  }
  \label{fig:measures-overview}
\end{figure}

Condition-level predictions are a probability distribution over response types, obtained from item-level scores $S(R_{l}, I_{k})$ by averaging over all items belonging to the relevant condition.
To map non-normalized scores $\myvec{s} = \tuple{s_{1}, \dots, s_{l}}$ to probabilities $\myvec{p} = \tuple{p_{1}, \dots, p_{l}}$ with different degrees of optimization, a common choice is the softmax function with optimality parameter $\alpha$:
%
\begin{align*}
 \text{SoftMax}(\myvec{s}, \alpha) = \myvec{p}, \text{where } p_{i} \propto \expo \left (\alpha p_{i} \right )\,.
\end{align*}
%
The softmax function can furthermore be decomposed into a first step of mapping scores to probabilities, and subsequently reweighing probabilities via a power-law transformation \citep[c.f.,][]{FrankeDegen2023:The-softmax-fun}:
%
\begin{align*}
  \text{SoftMax}(\myvec{s}, \alpha) & = \text{Pow}( \text{SoftMax}(\myvec{s}, 1) ; \alpha) \\
  \text{Pow}(\myvec{p} ; \alpha) & = \myvec{q}, \text{where } q_{i} \propto p_{i}^{\alpha}
\end{align*}
%
This means that there are three conceivable scope-sites for aggregating item-level information as shown in Figure~\ref{fig:measures-overview}.
Narrow-scope aggregation first aggregates the item-level scores, and then transposes the averages into (scaled) probabilities:
%
\begin{align*}
  % & \textbf{narrow-scope aggregation} \\
  & P_{n}(R_{l}, C ; \alpha) \propto \expo \left [  \alpha \ \frac{1}{m} \ \sum_{i = k}^{m} S(R_{l}, I_{k})  \right ]
    \tag*{\textcolor{gray}{[narrow-scope aggregation]}}
\end{align*}
%
Wide-scope aggregation, first transposes scores into probabilities, scales them and only aggregates over items last.
\begin{align*}
  % & \textbf{wide-scope aggregation} \\
  & P_{w}(R_{l}, C ; \alpha) \propto \frac{1}{m} \ \sum_{i = k}^{m} \frac{\expo \left( \alpha \ S(R_{l}, I_{k}) \right )}{ \sum_{l'} \expo \left( \alpha \ S(R_{l'}, I_{k}) \right )}
    \tag*{\textcolor{gray}{[wide-scope aggregation]}}
\end{align*}
%
Intermediate-scope aggregation performs item-level averaging after mapping scores onto probabilities (using softmax with $\alpha=1$), but before scaling (with a power-law transformation with variable $\alpha$):
\begin{align*}
  % & \textbf{intermediate-scope aggregation} \\
  & P_{i}(R_{l}, C ; \alpha) \propto  \left [ \frac{1}{m} \ \sum_{i = k}^{m} \frac{\expo \left( S(R_{l}, I_{k}) \right )}{ \sum_{l'} \expo \left( S(R_{l'}, I_{k}) \right )} \right ]^{\alpha}
    \tag*{\textcolor{gray}{[intermediate-scope aggregation]}}
\end{align*}

>>>>

The RSA model's predictions are defined at the condition-level.
An LLM, on the other hand, first and foremost makes predictions about each item.
As mentioned previously, the items of the reference game experiment differ in which levels of features (color, shape, texture) instantiate the structure of the task shown in Figure \ref{fig:ref-game}, as well as the order of presentation of objects and words.
We expect both human and LLM predictions to vary between different items: e.g., humans seem to have preferences for some features \citep[e.g.,][]{QingFranke2013:Variations-on-a}, such as over-production of informationally irrelevant material \citep[e.g.,][]{DaviesKatsos2010:Over-informativ,Rubio-Fernandez2019:Overinformative,DegenHawkins2020:When-redundancy}; and machines may be susceptible to the presentation of the order of choice options. \mf{@PT references from the NLP literature on effects of prompt variation?}
It therefore becomes an empirical question of whether a standard item-level score from LLMs provides a good fit, if we aim at predicting the human data separately for each item, not as a condition-level average.

As described in Section~\ref{motivation}, the standard procedure for assessing LLM predictions in multi-choice tasks is to use a winner-takes-all (WTA) strategy that just picks the best option for each item.
But since this strategy has conceptual and empirical weaknesses, we define the item-level LLM predictions using the more general softmax-approach, which contains the WTA-based approach as a limiting case, repeated here from above with slightly updated notation:
%
\begin{align*}
P_{\text{item}}^{\text{LLM}}\left ( y_{ki} \mid C; \alpha_{c} \right ) \propto \expo \left (\alpha_{c} \ \text{S}_{ki} \right)\,.
\end{align*}
%
Item-level scores $\text{S}_{ki}$, as defined in Section~\ref{motivation}, are obtained from GPT-3.5 \mf{which model exactly, when assessed, cite paper}.
% \footnote{
  % Numerically, we worked with \emph{average} log-probabilities. As in our experiment, all choice options had the same number of words, this is largely irrelevant, except for the effect of the $\alpha$ parameter in the interpretation condition. Using length-corrected log-probabilities is the same as using raw log-probabilities with an $\alpha$ parameter divided by the number of words. In this way, we make the $\alpha$ parameter more meaningfully comparable across conditions.}
The task description $x_{k}$ for item $I_{k}$ is a text supplied as prompt (see Appendix~\ref{sec:examples-items-llm} for an example).
The choice options are categorized as in the human experiment.\footnote{
  In the current set-up the response type ``distractor'' has two instantiations in the production condition. Since choice options are a single word in the production condition, for simplicity, we lump both of the distractor words together and treat them as a single option by taking the sum of the log-probabilities for both distractor words.}
Allowing for random errors with probability $\epsilon_{C}$, as before,:
%
\begin{align*}
  P_{\text{item}}^{\text{LLM}}\left ( y_{ki} \mid C; \alpha_{c}, \epsilon_{c} \right )
  = \epsilon_{c} \  P_{\text{item}}^{\text{LLM}}\left ( y_{ki} \mid C; \alpha_{c}, \epsilon_{c} \right ) \ \ + \ \ \nicefrac{\epsilon_{c}}{3}   \,.
\end{align*}
%
The resulting likelihood function for item-level data with item-level predictions is:
%
\begin{align*}
 P^{\text{LLM}}_{\text{item}}(D_{c} \mid C; \alpha_{c}, \epsilon_{c}) = \prod_{k = 1}^{m} \text{Multinomial} \left ( D_{c}^{k}, \tuple{P^{\text{LLM}}_{\text{item}}(y_{ki} \mid C;  \alpha_{c}, \epsilon_{c})}_{1 \le l \le 3} \right )\,.
\end{align*}
%
Priors on parameters are as for the RSA model described in Section~\ref{sec:model-pred-from}.

@@

Sampling-based approximations of Bayesian posterior predictive $p$-values for the by-item analysis are very low (see Table~\ref{tab:sumStats}), suggesting that the unaggregated item-level LLM predictions are inadequate predictors of the human data.
Figure~\ref{fig:item-level-obs-pred} shows that the item-level LLM scores predict variance which is not borne out by the human data.
Concretely, the plots show, for each condition and item, mean posterior estimates of the model's predicted probability of choosing the target option ($x$-axis), together with the observed proportion of target choices in the human data ($y$-axis).
There is ample variation in the model's predictions, especially visible in the production condition, owing to the fact that the item-level scores of the LLM sometimes clearly favor another option than the target choice.
So, the model itself predicts systematic variability at the item level.
The human data, too, show variability at the item-level, but there is no (visual) indication that the item-level variability predicted by the LLMs is borne out by the human data.
These results suggest that LLM-based probabilistic predictions may imply item-level variance that is not attested in the human data.
Put more strongly, a model that uses the most obvious item-level score to predict what human participants choose on a by-item level is ruled out by experimental data.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.9\textwidth]{00-pics/measures-overview.png}
  \caption{
    Schematic overview over three different approaches of obtaining condition-level predictions by aggregating item-level information.
    Approaches differ in the kind of item-level information they take into account and in the order in which aggregation, transformation to probabilities and parameterized scaling occur.
  }
  \label{fig:measures-overview}
\end{figure}

Condition-level predictions are a probability distribution over response types, obtained from item-level scores $\text{S}_{ki}$ by averaging over all items belonging to the relevant condition.
To map non-normalized scores $\myvec{s} = \tuple{s_{1}, \dots, s_{l}}$ to probabilities $\myvec{p} = \tuple{p_{1}, \dots, p_{l}}$ with different degrees of optimization, a common choice is the softmax function with optimality parameter $\alpha$:
%
\begin{align*}
 \text{SoftMax}(\myvec{s}, \alpha) = \myvec{p}, \text{where } p_{i} \propto \expo \left (\alpha p_{i} \right )\,.
\end{align*}
%
The softmax function can furthermore be decomposed into a first step of mapping scores to probabilities, and subsequently reweighing probabilities via a power-law transformation \citep[c.f.,][]{FrankeDegen2023:The-softmax-fun}:
%
\begin{align*}
  \text{SoftMax}(\myvec{s}, \alpha) & = \text{Pow}( \text{SoftMax}(\myvec{s}, 1) ; \alpha) \\
  \text{Pow}(\myvec{p} ; \alpha) & = \myvec{q}, \text{where } q_{i} \propto p_{i}^{\alpha}
\end{align*}
%
This means that there are three conceivable scope-sites for aggregating item-level information as shown in Figure~\ref{fig:measures-overview}.
Narrow-scope aggregation first aggregates the item-level scores, and then transposes the averages into (scaled) probabilities:
%
\begin{align*}
  & P_{\text{cond}}^{\text{SCR}}(r_{i} \mid C, \alpha_{c})
    \propto \expo \left [  \alpha_{c} \ \frac{1}{m} \ \sum_{k=1}^{m} \text{S}_{ki}  \right ]
    \tag*{\textcolor{gray}{[average scores (narrow-scope aggregation)]}}
\end{align*}
%
Wide-scope aggregation, first transposes scores into probabilities, scales them and only aggregates over items last.
\begin{align*}
  % & \textbf{wide-scope aggregation} \\
  & P_{\text{cond}}^{\text{PRB}}(r_{i} \mid C, \alpha_{c})
    = \frac{1}{m} \ \sum_{k=1}^{m} P_{\text{item}}^{\text{LLM}}(y_{ki} \mid C, \alpha_{c})
    \tag*{\textcolor{gray}{[average probabilities (wide-scope aggregation)]}}
\end{align*}
%
Intermediate-scope aggregation performs item-level averaging after mapping scores onto probabilities (using softmax with $\alpha=1$), but before scaling (with a power-law transformation with variable $\alpha$):
\begin{align*}
  % & \textbf{intermediate-scope aggregation} \\
  & P_{\text{cond}}^{\text{WTA}}(r_{i} \mid C, \alpha_{c})
    \propto  \left [ \frac{1}{m} \ \sum_{k=1}^{m} P_{\text{item}}^{\text{WTA}}(y_{ki} \mid C) \right ]^{\alpha_{c}}\,,
    \tag*{\textcolor{gray}{[average WTA (intermediate-scope aggregation)]}}
\end{align*}
where $P_{\text{item}}^{\text{WTA}}(y_{ki} \mid C) = \lim_{\alpha} P_{\text{item}}^{\text{LLM}}(y_{ki} \mid C, \alpha)$.
\bigskip
